---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rtrees

The goal of `rtrees` is to remove the barriers to derive phylogenies for a list of species from mega-trees. Basically, `Phylomatic` and more.


# Installation

```{r eval=FALSE}
options(repos = c(
  rtrees = 'https://daijiang.r-universe.dev',
  CRAN = 'https://cloud.r-project.org'))
install.packages("rtrees")
```

The above code will also install one dependency [{megatrees}](https://github.com/daijiang/megatrees), an R data package that hosts the megatrees used here.

## Shinny app

There is also a [Shinny app](https://djli.shinyapps.io/rtrees_shiny/) to get phylogenies without using R. However, given that I can only afford the free plan, the number of species to be inserted to the megatrees here will be limited to 1,000.

# Mega-trees

Currently, `rtrees` provides mega-trees for the following taxonomic groups: amphibian, bird, fish, mammal, plant, reptile, and shark_ray. The mega-trees (or subset of 100 posterior phylogenies; 50 for fish to reduce the package size) are saved in the data package [{megatrees}](https://github.com/daijiang/megatrees).

| Taxon                    | # of species | # of trees | R object (in `{megatrees}`)  | Reference            |
|--------------------------|--------------|------------|------------------------------|----------------------|
| Amphibian                | 7238         | 100        | `tree_amphibian_n100`        | Jetz and Pyron 2018  |
| Bee                      | 4651         | 1          | `tree_bee`                   | Henríquez-Piskulich et al. 2023 |
|                          | 4651         | 100        | `tree_bee_n100`              | Henríquez-Piskulich et al. 2023 |
| Butterfly                | 2244         | 1          | `tree_butterfly`             | Kawahara et al. 2023 |
| Bird                     | 9993         | 100        | `tree_bird_n100`             | Jetz et al. 2012     |
|                          | 11017        | 1          | `tree_bird_McTavish`         | McTavish et al. 2025     |
| Fish                     | 11638        | 1          | `tree_fish_12k`              | Rabosky et al. 2018  |
|                          | 31516        | 50         | `tree_fish_32k_n50`          | Rabosky et al. 2018  |
| Mammal                   | 5831         | 100        | `tree_mammal_n100_phylacine` | Faurby et al. 2018   |
|                          | 5911         | 100        | `tree_mammal_n100_vertlife`  | Upham et al. 2019    |
| Plant                    | 74531        | 1          | `tree_plant_otl`             | Smith and Brown 2018 |
| Reptile (Squamate)       | 9755         | 100        | `tree_reptile_n100`          | Tonini et al. 2016   |
| Shark, Ray, and Chimaera | 1192         | 100        | `tree_shark_ray_n100`        | Stein et al. 2018    |


# Usage examples


```{r}
library(rtrees)
library(ape)
```

## Prepare species data frame (optional)

The species lists which we want to have a phylogeny should be provided as a data frame with at least 3 columns: `family`, `genus`, and `species`. Their order does not matter. Here is an example for fish.

```{r echo = T}
test_fish_list = tibble::tibble(
  species = c("Serrasalmus_geryi", "Careproctus_reinhardti", "Gobiomorphus_coxii", 
              "Periophthalmus_barbarus", "Prognichthys_glaphyrae", "Barathronus_bicolor", 
              "Knipowitschia_croatica", "Rhamphochromis_lucius", "Neolissochilus_tweediei", 
              "Haplochromis_nyanzae", "Astronesthes_micropogon", "Sanopus_reticulatus"),
  genus = c("Serrasalmus", "Careproctus", "Gobiomorphus", "Periophthalmus",
            "Prognichthys", "Barathronus", "Knipowitschia", "Rhamphochromis", 
            "Neolissochilus", "Haplochromis", "Astronesthes", "Sanopus"),
  family = c("Serrasalmidae", "Liparidae", "Eleotridae", "Gobiidae", 
             "Exocoetidae", "Aphyonidae", "Gobiidae", "Cichlidae", 
             "Cyprinidae", "Cichlidae", "Stomiidae", "Batrachoididae")
)
```

```{r}
test_fish_list
```

For plant, fish, bird, and mammal, it is possible to prepare `sp_list` with function `sp_list_df()`; though you should check the results. It is possible to just provide the speices as a character vector for these taxon groups (or all genus in your species list are presented in the phylogeny).

```{r}
sp_list_df(sp_list = c("Periophthalmus_barbarus", "Barathronus_bicolor"),
           taxon = "fish")
```

## Get phylogeny from one megatree

Once we have the species list ready, we can then derive a phylogeny from `tree_fish`.

```{r}
test_tree = get_tree(sp_list = test_fish_list,
                     taxon = "fish",
                     scenario = "at_basal_node",
                     show_grafted = TRUE)
plot(test_tree, no.margin = T)
```

No matter whether `show_grafted` is 'TRUE' or 'FALSE', the grafting information was saved along with the phylogeny and can be extracted with the following code:

```{r echo=TRUE}
# or use rtrees::get_graft_status(test_tree)
test_tree$graft_status
```

## Get a set of phylogenies from multiple posterior megatrees

The function `rtrees::get_tree()` can also work with a set of posterior megatrees with the option to use parallel computing for the whole process. The default number of cores to be used will be the available number of cores minus 2 (so that users can still perform other tasks on their computers at the same time). The output will be a set of generated phylogenies with class `multiPhylo`; the number of derived phylogenies will be the same as the input megatrees. For this scenario, we can use exactly the same code described above. For example, if we want to use the 100 randomly selected posterior phylogenies that have ~32k fish species, we can add the `fish_tree = "all-taxon`.

```{r}
test_tree2 = get_tree(sp_list = test_fish_list,
                     taxon = "fish",
                     scenario = "at_basal_node",
                     show_grafted = TRUE,
                     fish_tree = "all-taxon")
test_tree2
plot(ladderize(test_tree2[[1]]), no.margin = T)
```

## Add tips to user provided trees

It is also possible to derive a phylogeny based on a user provided tree. This can be particularly useful if we have things like `Acer_spp` that we want to insert to the basal node of the genus.

```{r fig.width=9, fig.height=9}
test_tree = ape::read.tree(text = "(((((((((Potentilla_intermedia:1.156097,Potentilla_gracilis:1.156097):9.169741,Potentilla_bipinnatifida:10.325839):10.466967,Potentilla_sterilis:20.792806):11.333216,Fragaria_virginiana:32.126022):6.026567,(((((Rosa_setigera:3.434279,Rosa_arkansana:3.434279):1.991106,Rosa_spinosissima:5.425386):0.993924,Rosa_acicularis:6.419309):6.730804,(Rosa_micrantha:0.188519,Rosa_canina:0.188519):12.961594):23.677485,(Agrimonia_gryposepala:4.730863,Agrimonia_parviflora:4.730863):32.096735):1.324992):6.47604,(Geum_urbanum:11.153303,Geum_rivale:11.153303):33.475326):1.449852,((((((((Rubus_semisetosus:3.10529,Rubus_glandicaulis:3.10529):0.957584,Rubus_steelei:4.062873):0.398394,Rubus_cuneifolius:4.461267):2.271628,Rubus_vagus:6.732895):0.023439,Rubus_superioris:6.756333):0.020104,(Rubus_multifer:4.287114,Rubus_elegantulus:4.287115):2.489323):0.025475,((Rubus_laciniatus:0.125186,Rubus_bifrons:0.125186):0.11903,Rubus_praecox:0.244216):6.557697):11.571849,(Rubus_parviflorus:14.85915,Rubus_odoratus:14.85915):3.514612):27.704718):16.915234,Filipendula_rubra:62.993714):13.354116,((((((((((Crataegus_dissona:0.081863,Crataegus_florifera:0.081863):0.49838,Crataegus_fulleriana:0.580243):1.434261,Crataegus_crus-galli:2.014503):0.177737,Crataegus_chrysocarpa:2.192239):0.051159,Crataegus_uniflora:2.243399):0.438351,Crataegus_persimilis:2.68175):4.697259,((Malus_toringo:6.472694,Malus_ioensis:6.472693):0.772012,Chaenomeles_speciosa:7.244705):0.134305):1.72489,Pyrus_communis:9.1039):0.313393,Amelanchier_spicata:9.417293):40.039557,((((((Prunus_pumila:6.189944,Prunus_nigra:6.189945):0.37487,Prunus_pensylvanica:6.564815):4.856241,(Prunus_padus:4.816999,Prunus_serotina:4.817):6.604057):29.162527,Sorbaria_sorbifolia:40.583584):2.924475,Rhodotypos_scandens:43.50806):2.718224,(Spiraea_X_bumalda:25.921582,Aruncus_dioicus:25.921583):20.304701):3.230567):26.89098);")

test_tree_sp = c("Rosa_sp", "Rubus_sp", "Amelanchier_sp", "Prunus_sp", "Sorbus_sp", "Potentilla_sp",
  "Potentilla_intermedia", "Potentilla_gracilis", "Potentilla_bipinnatifida", "Potentilla_sterilis", "Fragaria_virginiana", "Rosa_setigera", "Rosa_arkansana", "Rosa_spinosissima", "Rosa_acicularis", "Rosa_micrantha", "Rosa_canina", "Agrimonia_gryposepala", "Agrimonia_parviflora", "Geum_urbanum", "Geum_rivale", "Rubus_semisetosus", "Rubus_glandicaulis", "Rubus_steelei", "Rubus_cuneifolius", "Rubus_vagus", "Rubus_superioris", "Rubus_multifer", "Rubus_elegantulus", "Rubus_laciniatus", "Rubus_bifrons", "Rubus_praecox", "Rubus_parviflorus", "Rubus_odoratus", "Filipendula_rubra", "Crataegus_dissona", "Crataegus_florifera", "Crataegus_fulleriana", "Crataegus_crus-galli", "Crataegus_chrysocarpa", "Crataegus_uniflora", "Crataegus_persimilis", "Malus_toringo", "Malus_ioensis", "Chaenomeles_speciosa", "Pyrus_communis", "Amelanchier_spicata", "Prunus_pumila", "Prunus_nigra", "Prunus_pensylvanica", "Prunus_padus", "Prunus_serotina", "Sorbaria_sorbifolia", "Rhodotypos_scandens", "Spiraea_X_bumalda", "Aruncus_dioicus")

plot(get_tree(sp_list = test_tree_sp, tree = test_tree, taxon = "plant",
              show_grafted = T, tree_by_user = T), type = "fan")
```

**Pay attention to the warning message there**. Because not all genera in the species list are presented in the megatree, it is impossible to find the most recent ancestor for genus and family that are missing from the megatree. Therefore, **for user-provided phylogenies, here is the recommended steps**:

1. Get the classification information (a data frame with two columns: genus and family) for all genera presented in the species list and the megatree.
2. Prepare the species list as a data frame with three columns: species, genus, and family using the classification information.
3. Process the megatree using function `processed_tree = rtrees::add_root_infor(tree, classification)`.
4. Pass the processed megatree to `rtrees::get_tree(sp_list_data_frame, processed_tree)` to derive the phylogeny for your species list.


## Bind missing species to specified places

It is also possible to specify a particular species to bind with by specifying columns `close_sp` and/or `close_genus`.

```{r fig.width=9, fig.height=9}
test_tree_sp_df = sp_list_df(test_tree_sp, "plant")
test_tree_sp_df$close_sp = NA
test_tree_sp_df$close_sp[1] = "Rosa_acicularis" # bind Rosa_sp to here
plot(get_tree(sp_list = test_tree_sp_df, tree = test_tree, taxon = "plant",
              show_grafted = T, tree_by_user = T), type = "fan")
```

# Some notes 

- If `tree` is specified, then `taxon` can be ignored if all genus in the species list are presented in the phylogeny.
- If a species does not have a co-family species in the mega-tree, it will be skipped.
- If `show_grafted = TRUE`, species that are grafted will have one or two `*` at the end of their names.
  + If it is grafted at the genus level, one `*`.
  + If it is grafted at the family level, two `*`s (i.e., `**`).
- The default scenario is `at_basal_node`, which will graft species at the genus/family basal node.
  + `random_below_basal` will randomly select a downstream node to attach the new tip.
  + If only one species in the mega-tree that is in the same genus/family of the new tip, then the new tip will be grafted at the middle of this species' branch for all scenarios.
- The `tree` can be a user provided tree, if so, set `tree_by_user = TRUE`.
- See `?get_tree` for more details.
  
# Contribution

Feel free to test it. Contributions and suggestions are welcome. You can open an issue or send a pull request.
  
# Citation

Li Daijiang. (2023). rtrees: an R package to assemble phylogenetic trees from megatrees. Ecography.


