---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# rtrees

The goal of `rtrees` is to remove the barriers to derive phylogenies for a list of species from mega-trees. Basically, `Phylomatic` and more.


# Installation

It may take a while because one of its dependency [{megatrees}](https://github.com/daijiang/megatrees) is relatively large.

```{r eval=FALSE}
devtools::install_github("daijiang/rtrees")
```

```{r}
library(rtrees)
library(ape)
```

# Mega-trees

Currently, `rtrees` provides mega-trees for the following taxonomic groups: amphibian, bird, fish, mammal, plant, reptile, and shark_ray. The mega-trees (or subset of 100 posterior phylogenies) are saved in the data package [{megatrees}](https://github.com/daijiang/megatrees), which takes a while to install because of its relatively large size.

| Taxon                    | # of species | # of trees | R object (in `{megatrees}`)  | Reference            |
|--------------------------|--------------|------------|------------------------------|----------------------|
| Amphibian                | 7238         | 100        | `tree_amphibian_n100`        | Jetz and Pyron 2018  |
| Bird                     | 9993         | 100        | `tree_bird_n100`             | Jetz et al. 2012     |
| Fish                     | 11638        | 1          | `tree_fish_12k`              | Rabosky et al. 2018  |
|                          | 31516        | 50         | `tree_fish_32k_n50`          | Rabosky et al. 2018  |
| Mammal                   | 5831         | 100        | `tree_mammal_n100_phylacine` | Faurby et al. 2018   |
|                          | 5911         | 100        | `tree_mammal_n100_vertlife`  | Upham et al. 2019    |
| Plant                    | 74531        | 1          | `tree_plant_otl`             | Brown and Smith 2018 |
| Reptile (Squamate)       | 9755         | 100        | `tree_reptile_n100`          | Tonini et al. 2016   |
| Shark, Ray, and Chimaera | 1192         | 100        | `tree_shark_ray_n100`        | Stein et al. 2018    |


# Usage examples

The species lists which we want to have a phylogeny should be provided as a data frame with at least 3 columns: `family`, `genus`, and `species`. Their order does not matter. Here is an example for fish.

```{r echo = T}
test_fish_list = tibble::tibble(
  species = c("Serrasalmus_geryi", "Careproctus_reinhardti", "Gobiomorphus_coxii", 
              "Periophthalmus_barbarus", "Prognichthys_glaphyrae", "Barathronus_bicolor", 
              "Knipowitschia_croatica", "Rhamphochromis_lucius", "Neolissochilus_tweediei", 
              "Haplochromis_nyanzae", "Astronesthes_micropogon", "Sanopus_reticulatus"),
  genus = c("Serrasalmus", "Careproctus", "Gobiomorphus", "Periophthalmus",
            "Prognichthys", "Barathronus", "Knipowitschia", "Rhamphochromis", 
            "Neolissochilus", "Haplochromis", "Astronesthes", "Sanopus"),
  family = c("Serrasalmidae", "Liparidae", "Eleotridae", "Gobiidae", 
             "Exocoetidae", "Aphyonidae", "Gobiidae", "Cichlidae", 
             "Cyprinidae", "Cichlidae", "Stomiidae", "Batrachoididae")
)
```

```{r}
test_fish_list
```

For plant, fish, bird, and mammal, it is possible to prepare `sp_list` with function `sp_list_df()`; though you should check the results. It is possible to just provide the speices as a character vector for these taxon groups (or all genus in your species list are presented in the phylogeny).

```{r}
sp_list_df(sp_list = c("Periophthalmus_barbarus", "Barathronus_bicolor"),
           taxon = "fish")
```


Then we can derive a phylogeny from `tree_fish`.

```{r}
test_tree = get_tree(sp_list = test_fish_list,
                     taxon = "fish",
                     scenario = "at_basal_node",
                     show_grafted = TRUE)
plot(test_tree, no.margin = T)
```

If we want to use the 100 randomly selected posterior phylogenies that have ~32k fish species, we can add the `fish_tree = "all-taxon`.

```{r}
test_tree2 = get_tree(sp_list = test_fish_list,
                     taxon = "fish",
                     scenario = "at_basal_node",
                     show_grafted = TRUE,
                     fish_tree = "all-taxon")
test_tree2
plot(ladderize(test_tree2[[1]]), no.margin = T)
```

  
## Add tips to user provided trees

It is also possible to derive a phylogeny based on a user provided tree. This can be particularly useful if we have things like `Acer_spp` that we want to insert to the basal node of the genus.

```{r fig.width=9, fig.height=9}
test_tree = ape::read.tree(text = "(((((((((Potentilla_intermedia:1.156097,Potentilla_gracilis:1.156097):9.169741,Potentilla_bipinnatifida:10.325839):10.466967,Potentilla_sterilis:20.792806):11.333216,Fragaria_virginiana:32.126022):6.026567,(((((Rosa_setigera:3.434279,Rosa_arkansana:3.434279):1.991106,Rosa_spinosissima:5.425386):0.993924,Rosa_acicularis:6.419309):6.730804,(Rosa_micrantha:0.188519,Rosa_canina:0.188519):12.961594):23.677485,(Agrimonia_gryposepala:4.730863,Agrimonia_parviflora:4.730863):32.096735):1.324992):6.47604,(Geum_urbanum:11.153303,Geum_rivale:11.153303):33.475326):1.449852,((((((((Rubus_semisetosus:3.10529,Rubus_glandicaulis:3.10529):0.957584,Rubus_steelei:4.062873):0.398394,Rubus_cuneifolius:4.461267):2.271628,Rubus_vagus:6.732895):0.023439,Rubus_superioris:6.756333):0.020104,(Rubus_multifer:4.287114,Rubus_elegantulus:4.287115):2.489323):0.025475,((Rubus_laciniatus:0.125186,Rubus_bifrons:0.125186):0.11903,Rubus_praecox:0.244216):6.557697):11.571849,(Rubus_parviflorus:14.85915,Rubus_odoratus:14.85915):3.514612):27.704718):16.915234,Filipendula_rubra:62.993714):13.354116,((((((((((Crataegus_dissona:0.081863,Crataegus_florifera:0.081863):0.49838,Crataegus_fulleriana:0.580243):1.434261,Crataegus_crus-galli:2.014503):0.177737,Crataegus_chrysocarpa:2.192239):0.051159,Crataegus_uniflora:2.243399):0.438351,Crataegus_persimilis:2.68175):4.697259,((Malus_toringo:6.472694,Malus_ioensis:6.472693):0.772012,Chaenomeles_speciosa:7.244705):0.134305):1.72489,Pyrus_communis:9.1039):0.313393,Amelanchier_spicata:9.417293):40.039557,((((((Prunus_pumila:6.189944,Prunus_nigra:6.189945):0.37487,Prunus_pensylvanica:6.564815):4.856241,(Prunus_padus:4.816999,Prunus_serotina:4.817):6.604057):29.162527,Sorbaria_sorbifolia:40.583584):2.924475,Rhodotypos_scandens:43.50806):2.718224,(Spiraea_X_bumalda:25.921582,Aruncus_dioicus:25.921583):20.304701):3.230567):26.89098);")

test_tree_sp = c("Rosa_sp", "Rubus_sp", "Amelanchier_sp", "Prunus_sp", "Sorbus_sp", "Potentilla_sp",
  "Potentilla_intermedia", "Potentilla_gracilis", "Potentilla_bipinnatifida", "Potentilla_sterilis", "Fragaria_virginiana", "Rosa_setigera", "Rosa_arkansana", "Rosa_spinosissima", "Rosa_acicularis", "Rosa_micrantha", "Rosa_canina", "Agrimonia_gryposepala", "Agrimonia_parviflora", "Geum_urbanum", "Geum_rivale", "Rubus_semisetosus", "Rubus_glandicaulis", "Rubus_steelei", "Rubus_cuneifolius", "Rubus_vagus", "Rubus_superioris", "Rubus_multifer", "Rubus_elegantulus", "Rubus_laciniatus", "Rubus_bifrons", "Rubus_praecox", "Rubus_parviflorus", "Rubus_odoratus", "Filipendula_rubra", "Crataegus_dissona", "Crataegus_florifera", "Crataegus_fulleriana", "Crataegus_crus-galli", "Crataegus_chrysocarpa", "Crataegus_uniflora", "Crataegus_persimilis", "Malus_toringo", "Malus_ioensis", "Chaenomeles_speciosa", "Pyrus_communis", "Amelanchier_spicata", "Prunus_pumila", "Prunus_nigra", "Prunus_pensylvanica", "Prunus_padus", "Prunus_serotina", "Sorbaria_sorbifolia", "Rhodotypos_scandens", "Spiraea_X_bumalda", "Aruncus_dioicus")

plot(get_tree(sp_list = test_tree_sp, tree = test_tree, taxon = "plant",
              show_grafted = T, tree_by_user = T), type = "fan")
```

It is also possible to specify a particular species to bind with by specifying columns `close_sp` and/or `close_genus`.

```{r fig.width=9, fig.height=9}
test_tree_sp_df = sp_list_df(test_tree_sp, "plant")
test_tree_sp_df$close_sp = NA
test_tree_sp_df$close_sp[1] = "Rosa_acicularis" # bind Rosa_sp to here
plot(get_tree(sp_list = test_tree_sp_df, tree = test_tree, taxon = "plant",
              show_grafted = T, tree_by_user = T), type = "fan")
```

Some notes: 

- If `tree` is specified, then `taxon` can be ignored if all genus in the species list are presented in the phylogeny.
- If a species does not have a co-family species in the mega-tree, it will be skipped.
- If `show_grafted = TRUE`, species that are grafted will have one or two `*` at the end of their names.
  + If it is grafted at the genus level, one `*`.
  + If it is grafted at the family level, two `*`s.
- The default scenario is `at_basal_node`, which will graft species at the genus/family basal node.
  + `random_below_basal` will randomly select a downstream node to attach the new tip.
  + `at_or_above_basal` will graft the new tip above the genus/family basal node if no co-family species found.
  + If only one species in the mega-tree that is in the same genus/family of the new tip, then the new tip will be grafted at the middle of this species' branch for all scenarios.
- The `tree` can be a user provided tree, if so, set `tree_by_user = TRUE`.
- See `?get_tree` for more details.
  
## Contribution

Feel free to test it. Contributions and suggestions are welcome. You can open an issue or send a pull request.
  

